Wait-free Queues with Polygarithmic Step Complexity

### 2022-06-22 article/psuedocode.tex
review pages 1-2

\newcommand{\sub}[1]{\textsubscript{#1}}
\renewcommand{\tt}[1]{\texttt{#1}}
\renewcommand{\sl}[1]{\textsl{#1}}
\renewcommand{\it}[1]{\textit{#1}}
\renewcommand{\bf}[1]{\textbf{#1}}
\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
\newcommand{\cmt}[1]{\Comment{#1}}

@@

\Statex $\diamondsuit$ \tt{\sl{Shared}}
\begin{itemize}
\item \tt{\sl{Tree} tree} \textsf{: A binary tree of \tt{Node}s. \tt{root} is a pointer to the root node.}
%such that \tt{root} is the root and the left child and the right child and the parent of \tt{Tree[i]} are \tt{Tree[2i+1]}, \tt{Tree[2i+2]} and \tt{Tree[i/2]}.
\end{itemize}

@@

\Statex $\diamondsuit$ \tt{\sl{Local}}
\begin{itemize}
\item \tt{\sl{*Node} leaf:} \sf{ a pointer to the process's leaf in the tree.}
\end{itemize}

@@

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
\item \tt{\sl{BlockList} blocks}
  \textsf{ implemented with an array.}
\item \tt{\sl{int} size= 1}\textsf{: \#\tt{block}s in \tt{blocks}.}
\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent. Since it is incremented after propagating, it may be behind by 1.}
\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores an approximate index of the superblock of the blocks in \tt{blocks} whose \tt{group} field have value \tt{i}.}
\end{itemize}

@@

\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
\begin{itemize}
  \item \tt{\sl{Object} element}
  \textsf{: Each block in a leaf represents a single operation. For \tt{enqueue} operations element is the input of the \tt{enqueue} and for \tt{dequeue} operations it is \tt{null}.}

    \item \tt{\sl{Object} response}
  \textsf{: stores the response of the operation in the \tt{LeafBlock}.}

@@

\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
\begin{itemize}
  \item \tt{\sl{int} length}
  \textsf{: length of the queue after performing all operations in the prefix for this block}
%  \item \tt{\sl{int} sum\sub{non-null deq}}
%  \textsf{: count of non-null dequeus up to this block}
%  \item \tt{\sl{counter} num\sub{finished}}
%  \textsf{: number of finished operations in the block}
%    \item \tt{\sl{int} order}
%  \textsf{: the index of the block in the \tt{BlockList} containing the block.}
\end{itemize}

@@

\Statex
\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
\Statex \tt{blocks[]: }\sf{array of blocks}
\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
\State \Return{\tt{CAS(blocks[n], null, blk)}}
\EndFunction{TryAppend}

@@

\it{Variable naming:}
\begin{itemize}
  \item \tt{b\sub{op}}: index of the block containing operation \tt{op}
  \item \tt{r\sub{op}}: rank of operation \tt{op} i.e. the ordering among the operations of its type according to linearization ordering
\end{itemize}

@@

\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a block with element \tt{e} and appends it to the tree.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
\State \tt{newBlock.element= e}
%\State \tt{b.num\sub{enq}=1}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}+1}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}}
\State \tt{leaf.}\Call{Append}{newBlock}
\EndFunction{Enqueue}

@@

\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Computes the rank  and index of the block in the root of the enqueue that is the response of the \tt{i}th dequeue in the root's \tt{b}th block. Returns \tt{<-1,-->} if the queue is empty.}
\If{\tt{root.blocks[b-1].length + root.blocks[b].num\sub{enq} - i $<$ 0}}
\State \Return \tt{<-1,-->}
\Else
\Statex \cmt{We call the dequeues that return a value \it{non-null dequeues}. $r$th non-null dequeue returns the element of th $r$th enqueue. We can compute \# non-null dequeues in the prefix for a block this way: \#non-null dequeues= length - \#enqueues. Note that the $i$th dequeue in the given block is not a non-null dequeue.}
\State \tt{r\sub{enq}= root.blocks[b-1].sum\sub{enq}- root.blocks[b-1].length + i}
\State \Return \tt{<root.\Call{DSearch}{r\sub{enq}, b}, r\sub{enq}>}
\EndIf
\EndFunction{FindResponse}

@@

\Function{Object}{Dequeue()}{}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.size, 1}\cmt{\tt{r} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
\State \tt{<b\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{b\sub{deq}, r\sub{deq}} 
 \cmt{\tt{r\sub{enq}} is the rank of the enqueue whose element is the response to the dequeue in the block containing it and \tt{b\sub{deq}} is the index of that block of it in the blocklist. If the response is \tt{null} then r\sub{deq} is -1.} \label{deqRest}
\If{\tt{r\sub{enq}==-1}}
\State \tt{output= null}
\pagebreak
\Else
\State \tt{output= \Call{GetEnq}{b\sub{enq}, r\sub{enq}}} \cmt{getting the reponse's \tt{element}.}
\EndIf
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\Function{boolean}{Refresh()}{}
\State \tt{s= size} \label{readSize}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{s}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} field.}
\If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
\ElsIf{\tt{blocks.tryAppend(new, s)}} \label{cas}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.}
\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
\EndFor
\State \tt{\Call{CAS}{size, s, s+1}} \label{incrementHead1}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{size, s, s+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.} \label{incrementHead2}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{r\sub{enq}}}}
\Function{int}{DSearch}{\sl{int} i, \sl{int} end}

\Statex \cmt{\textmd{Searches for~the \tt{i}th enqueue of the given prefix sum of \tt{b}th block in the \tt{root}. Returns the index of the leftmost block in \tt{root.blocks} whose \sl{sum\sub{enq}} is $\geq$ \tt{i}}.}
\State \tt{start= b-1}
\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{r\sub{enq}}}
\State \tt{start= start-(b-start)}
\EndWhile
\Return \tt{root.BSearch(sum\sub{enq}, r\sub{enq}, start, b)}
\EndFunction{DSearch}

@@

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\Statex\cmt{Creates a block to be inserted into as $i$th \tt{block} in \tt{blocks}. Returns the created block as well as values read from each child's num\sub{propagated} field. These values are used for incrementing the children's num\sub{propagated} field if the block was appended to \tt{blocks} successfully.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\State \tt{newBlock.group= num\sub{propagated}}
\State \tt{newBlock.order= i}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.size} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{np\sub{dir}= dir.num\sub{propagated}}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
%\State \tt{b.sum= n.blocks[i-1].sum + b.num}

\If{\tt{this} \bf{is} \tt{root}}
\State \tt{newBlock.length= max(root.blocks[i-1].length + b.num\sub{enq} - b.num\sub{deq}, 0)}\label{computeLength}
%\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
\EndIf

@@

\Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
\State \tt{size+=1} \label{appendEnd} 
\State \tt{blk.group= size} \label{appendStart} 
\State \tt{blocks[size]= blk} 
\State \tt{parent.}\Call{Propagate()}{} 
\EndFunction{Append}

%\Statex
%
%\Function{void}{Help}{}\cmt{Helps pending operations}
%
%\State{\tt{last= l.size-1}}
%\cmt{\tt{l.blocks[last]} can not be \tt{null} because size increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
%\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
%\State \tt{l.blocks[last].response= l.}\Call{HelpDequeue()}{}
%\EndIf
%
%\EndFunction{Help}

@@

\framebox[1.1\width]{TEST} As a temporary test I have changed the name of n.size to n.head here, other options are n.head and n.lastBLock but they might be confusing since we have used them before. Fix the logical order of definitions (cyclic refrences).

@@

\begin{lemma}[headProgress] \label{lem::sizeProgress}
\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left}} $\geq$ \nf{n.blocks[i-1].end\sub{left}},\nf{n.blocks[i].end\sub{right}} $\geq$ \nf{n.blocks[i-1].end\sub{right}}.
\end{lemma}
\begin{proof}
The first claim follows trivially from the pseudocode since \nf{n.head} is only incremented. Also  when \nf{n.blocks[i]} is created its \nf{end\sub{left}, end\sub{right}} are greater than or equal to the values in \nf{n.blocks[i-1]}. Since \nf{blocks[i-1].end\sub{dir}}$<$\nf{dir.size = blocks[i].end\sub{dir}} (Lines~\ref{lastLine, prevLine}).
\end{proof}

@@

Some value is written into \nf{n.blocks[head]} only in Line 313. It is obvious that writing into \nf{n.blocks[head]} presereves the invariant. The value of \nf{n.head} is modified only in lines \ref{incrementHead1}], \ref{incrementHead2}. Depending on wether the \nf{TryAppend()}in Line\ref{cas} succeeded or not we show that the claim holds after the increment lines of \nf{n.head} in either case. If \nf{head} is incremented to $h$ it is sufficient to show \nf{n.blocks[h]}$\neq$\nf{null} to prove the invariant still holds. In the first case the process applied a successful \nf{TryAppend(new,h)} in line \ref{okcas}, which means \nf{n.blocks[h]} is not null anymore. Note that wether\ref{incrementHead1} returns true or false after Line \nf{n.head} we know has been incremented from Line \ref{readSize}. The failure case is also the same since it means some value is written into \nf{n.blocks[head]} by some process.
%At the end of every \nf{n.Refresh()} with a block  with size greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}). If a process went to sleep before incrementing the \nf{established} (line \ref{incrementHead1}), nothing can be appended to \nf{n.blocks} until another process increments \nf{n.establsihed} (Line \ref{incrementHead2}).
\end{proof}
\it{Explain More}

@@

%\begin{lemma}[createBlock] \label{lem::createBlock}
%  If \nf{n.CreateBlock(i)} is invoked at time $t$ and returns block $new$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(new)}.
%\end{lemma}
%\begin{proof}
%From lemmas 7,8 we know that \nf{n.blocks[i-1]} is full and assume it is invoked at $t^\prime$. 
%We want to prove every block $b$ that is $\in$ \nf{EST\sub{n.left,t}} $\cup$ \nf{EST\sub{n.right,t}} is also $\in$ \nf{EST\sub{n,t}}$^\prime$ $\cup$ \nf{new}. It is easy to see \nf{EST\sub{n.left,t}} $\cap$ \nf{EST\sub{n.right,t}} is $\emptyset$. If $b$ is $\in$ \nf{EST\sub{n.left,t$^\prime$}} it is simple to see. Else we can see from the code of the \nf{CreateBlock} that \nf{new} contains it. Right child is the same.
%% WLOG we prove the claim for the left child. Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are established at time $t$.  Since the head is only increasing (Lemma~\ref{lem::sizeProgress}) the  claim holds after $t$. See Figure~\ref{fig::createBlock}. The right child is the same.
%\end{proof}

@@

%Suppose the previous true returning \nf{n.Refresh()} took place in time $(t_i^\prime-t_t^\prime)$. By induction on the number of succsuful \nf{Refresh()}'s \nf{n} we know that \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})}. Operations that are in \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})} but not in \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})} are in the block returned by \nf{n.CreateBlock()} by lemma 10. 
%By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readSize} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
\end{proof}
\it{Mention this new block is establish in t\sub{t}. last sentece is not complete}

>>>>

\newcommand{\sub}[1]{\textsubscript{#1}}
\renewcommand{\tt}[1]{\texttt{#1}}
\renewcommand{\sl}[1]{\textsl{#1}}
\renewcommand{\it}[1]{\textit{#1}}
\renewcommand{\bf}[1]{\textbf{#1}}
\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
\newcommand{\cmt}[1]{\Comment{#1}}
\newcommand{\head}{head}
\newcommand{\size}{size}

@@

\Statex $\diamondsuit$ \tt{\sl{Shared}}
\begin{itemize}
\item \tt{\sl{Tree} tree} \textsf{: A binary tree of \tt{Node}s. \tt{root} is the root node.}
%such that \tt{root} is the root and the left child and the right child and the parent of \tt{Tree[i]} are \tt{Tree[2i+1]}, \tt{Tree[2i+2]} and \tt{Tree[i/2]}.
\end{itemize}

@@

\Statex $\diamondsuit$ \tt{\sl{Local}}
\begin{itemize}
\item \tt{\sl{Node} leaf:} \sf{ process's leaf in the tree.}
\end{itemize}

@@

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
\item \tt{\sl{BlockList} blocks}
  \textsf{ implemented with an array.}
\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}(-1). \tt{blocks[0]} is a block with all fields equal to zero.}
\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent. Since it is incremented after propagating, it may be behind by 1.}
\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores an approximate index of the superblock of the blocks in \tt{blocks} whose \tt{group} field have value \tt{i}.}
\end{itemize}

@@

\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
\begin{itemize}
  \item \tt{\sl{Object} element}
  \textsf{: Each block in a leaf represents a single operation. If the operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}

@@

\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
\begin{itemize}
  \item \tt{\sl{int} \size}
  \textsf{: size of the queue after performing all operations in the prefix for this block}
%  \item \tt{\sl{int} sum\sub{non-null deq}}
%  \textsf{: count of non-null dequeus up to this block}
%  \item \tt{\sl{counter} num\sub{finished}}
%  \textsf{: number of finished operations in the block}
%    \item \tt{\sl{int} order}
%  \textsf{: the index of the block in the \tt{BlockList} containing the block.}
\end{itemize}

@@

@@

%\it{Variable naming:}
%\begin{itemize}
%  \item \tt{b\sub{op}}: index of the block containing operation \tt{op}
%  \item \tt{r\sub{op}}: rank of operation \tt{op} i.e. the ordering among the operations of its type according to linearization ordering
%\end{itemize}

@@

\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a block with element \tt{e} and appends it to the tree.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
\State \tt{newBlock.element= e}
%\State \tt{b.num\sub{enq}=1}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
\State \tt{leaf.head+=1}
\State \tt{leaf.}\Call{Append}{newBlock}
\EndFunction{Enqueue}

@@

\Function{<int, int>}{FindResponse}{\sl{int} b\sub{d}, \sl{int} i\sub{d}}\cmt{If $E(root,b_e,i_e)$ is the response to the $D(root,b_d,i_d)$ returns \tt{<b\sub{e},i\sub{e}>}. Returns \tt{<-1,-->} if the queue is empty.}
\If{\tt{ root.blocks[b\sub{d}-1].\size}\tt{ + root.blocks[b\sub{d}].num\sub{enq} - (i + root.blocks[b\sub{d}-1].sum\sub{deq}) $<$ 0}}
\State \Return \tt{<-1,-->}
\Else

\State \tt{r\sub{enq}= i + root.blocks[b\sub{d}-1].sum\sub{deq} - (root.blocks[b\sub{d}-1].size - root.blocks[b\sub{d}-1].sum\sub{enq} + root.blocks[b\sub{d}-1].sum\sub{deq})}
\State \cmt{size-enqs+deqs=null deqs}
\State \Return \tt{root.\Call{DSearch}{r\sub{enq}, b\sub{d}}}
\EndIf
\EndFunction{FindResponse}

@@

\Function{Object}{Dequeue()}{}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.\head, 1}\cmt{\tt{r\sub{deq}} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
\State \tt{<b\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{b\sub{deq}, r\sub{deq}} 
 \cmt{\tt{$E(root,b_{enq},i_{enq})$ is response to the $D(root,b_{deq},i_{deq})$. If the response is \tt{null} then r\sub{enq} is -1.}} \label{deqRest}
\If{\tt{r\sub{enq}==-1}}
\State \tt{output= null}
\pagebreak
\Else
\State \tt{output= \Call{GetEnq}{b\sub{enq}, r\sub{enq}}} \cmt{getting the reponse's \tt{element}.}
\EndIf
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\Function{boolean}{Refresh()}{}
\State \tt{s= \head} \label{readHead}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{s}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} field.}
\If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
\ElsIf{\tt{blocks.tryAppend(new, s)}} \label{cas}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.}
\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
\EndFor
\State \tt{\Call{CAS}{\head, s, s+1}} \label{incrementHead1}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{\head, s, s+1}} \cmt{Even if another process wins, help to increase the \head. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{r\sub{enq}}}}
\Function{<int, int>}{DSearch}{\sl{int} i, \sl{int} end}

\Statex \cmt{\textmd{Searches for~the \tt{i}th enqueue of the given prefix sum of \tt{b}th block in the \tt{root}. Returns the index of the leftmost block in \tt{root.blocks} whose \sl{sum\sub{enq}} is $\geq$ \tt{i}}.}
\State \tt{start= b-1}
\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{r\sub{enq}}}
\State \tt{start= start-(b-start)}
\EndWhile \State
\Return \tt{root.BSearch(sum\sub{enq}, r\sub{enq}, start, b)}
\EndFunction{DSearch}

@@

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\Statex\cmt{Creates a block to be inserted into as $i$th \tt{block} in \tt{blocks}. Returns the created block as well as values read from each child's num\sub{propagated} field. These values are used for incrementing the children's num\sub{propagated} field if the block was appended to \tt{blocks} successfully.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\State \tt{newBlock.group= num\sub{propagated}}
\State \tt{newBlock.order= i}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.\head} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{np\sub{dir}= dir.num\sub{propagated}}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
%\State \tt{b.sum= n.blocks[i-1].sum + b.num}

\If{\tt{this} \bf{is} \tt{root}}
\State \tt{newBlock.\size= max(root.blocks[i-1].\size + b.num\sub{enq} - b.num\sub{deq}, 0)}\label{computeLength}
%\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
\EndIf

@@

\Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
\State \tt{\head+=1} \label{appendEnd} 
\State \tt{blk.group= \head} \label{appendStart} 
\State \tt{blocks[\head]= blk} 
\State \tt{parent.}\Call{Propagate()}{} 
\EndFunction{Append}

%\Statex
%
%\Function{void}{Help}{}\cmt{Helps pending operations}
%
%\State{\tt{last= l.\head-1}}
%\cmt{\tt{l.blocks[last]} can not be \tt{null} because \head increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
%\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
%\State \tt{l.blocks[last].response= l.}\Call{HelpDequeue()}{}
%\EndIf
%
%\EndFunction{Help}

\Statex
%\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
\Statex \tt{blocks[]: }\sf{array of blocks}
\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
\State \Return{\tt{CAS(blocks[n], null, blk)}}
\EndFunction{TryAppend}

@@

\framebox[1.1\width]{TEST} As a temporary test I have changed the name of n.\head to n.head here, other options are n.head and n.lastBLock but they might be confusing since we have used them before. Fix the logical order of definitions (cyclic refrences).

@@

\begin{lemma}[headProgress] \label{lem::headProgress}
\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left}} $\geq$ \nf{n.blocks[i-1].end\sub{left}},\nf{n.blocks[i].end\sub{right}} $\geq$ \nf{n.blocks[i-1].end\sub{right}}.
\end{lemma}
\begin{proof}
The first claim follows trivially from the pseudocode since \nf{n.head} is only incremented. Also  when \nf{n.blocks[i]} is created its \nf{end\sub{left}, end\sub{right}} are greater than or equal to the values in \nf{n.blocks[i-1]}. Since \nf{blocks[i-1].end\sub{dir}}$<$\nf{dir.\head = blocks[i].end\sub{dir}} (Lines~\ref{lastLine, prevLine}).
\end{proof}

@@

Some value is written into \nf{n.blocks[head]} only in Line 313. It is obvious that writing into \nf{n.blocks[head]} presereves the invariant. The value of \nf{n.head} is modified only in lines \ref{incrementHead1}], \ref{incrementHead2}. Depending on wether the \nf{TryAppend()}in Line\ref{cas} succeeded or not we show that the claim holds after the increment lines of \nf{n.head} in either case. If \nf{head} is incremented to $h$ it is sufficient to show \nf{n.blocks[h]}$\neq$\nf{null} to prove the invariant still holds. In the first case the process applied a successful \nf{TryAppend(new,h)} in line \ref{okcas}, which means \nf{n.blocks[h]} is not null anymore. Note that wether\ref{incrementHead1} returns true or false after Line \nf{n.head} we know has been incremented from Line \ref{readHead}. The failure case is also the same since it means some value is written into \nf{n.blocks[head]} by some process.
%At the end of every \nf{n.Refresh()} with a block  with \head greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}). If a process went to sleep before incrementing the \nf{established} (line \ref{incrementHead1}), nothing can be appended to \nf{n.blocks} until another process increments \nf{n.establsihed} (Line \ref{incrementHead2}).
\end{proof}
\it{Explain More}

@@

%\begin{lemma}[createBlock] \label{lem::createBlock}
%  If \nf{n.CreateBlock(i)} is invoked at time $t$ and returns block $new$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(new)}.
%\end{lemma}
%\begin{proof}
%From lemmas 7,8 we know that \nf{n.blocks[i-1]} is full and assume it is invoked at $t^\prime$. 
%We want to prove every block $b$ that is $\in$ \nf{EST\sub{n.left,t}} $\cup$ \nf{EST\sub{n.right,t}} is also $\in$ \nf{EST\sub{n,t}}$^\prime$ $\cup$ \nf{new}. It is easy to see \nf{EST\sub{n.left,t}} $\cap$ \nf{EST\sub{n.right,t}} is $\emptyset$. If $b$ is $\in$ \nf{EST\sub{n.left,t$^\prime$}} it is simple to see. Else we can see from the code of the \nf{CreateBlock} that \nf{new} contains it. Right child is the same.
%% WLOG we prove the claim for the left child. Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are established at time $t$.  Since the head is only increasing (Lemma~\ref{lem::headProgress}) the  claim holds after $t$. See Figure~\ref{fig::createBlock}. The right child is the same.
%\end{proof}

@@

%Suppose the previous true returning \nf{n.Refresh()} took place in time $(t_i^\prime-t_t^\prime)$. By induction on the number of succsuful \nf{Refresh()}'s \nf{n} we know that \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})}. Operations that are in \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})} but not in \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})} are in the block returned by \nf{n.CreateBlock()} by lemma 10. 
%By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readHead} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
\end{proof}
\it{Mention this new block is establish in t\sub{t}. last sentece is not complete}

%\begin{lemma}[falseRefresh] \label{lem:falseRefresh}
%    If instance \nf{r} of \nf{n.Refresh()} reads value \nf{s} on line \ref{readHead} and then returns \nf{false}, then there is another instance $r^\prime$ of \nf{n.Refresh()} that has performed a successful \nf{TryAppend(new, s)}. A \nf{TryAppend()} is successful if its \nf{CAS} is successful.
%\end{lemma}
%\begin{proof}
%  If there is no other concurrent successful \nf{n.Refresh()} then \nf{n.Refresh()} would succeed in Line 313. So there is another \nf{n.Refresh()}, that has appended its block in \nf{n.blocks[h]} after Line 310 of the first \nf{n.Refresh()}. 
%%Otherwise the other \nf{Refresh(n)} should have read \nf{h$^\prime$>h} instead of \nf{h} for \nf{n.head}(Line 52).
%\end{proof}
